Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
edits sections 5 & 6

\begin{figure}[t!]
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/unexplained_effect.pdf}
		\caption{}
	\end{subfigure}%
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/ll.pdf}
		\caption{}
	\end{subfigure}
	\caption{Model fit improvement over iterations on the Adult Income dataset. In the expert simulation, an 
		LLM is used to determine the edge orientations. (a) Shows the total residual association, $\tau$. 
		(b) Shows the log-likelihood of the overall model.
	}
	\label{fig:unexplained_ll}
\end{figure}

@@

We now present two proof-of-concept implementations of our approach, 
geared towards using  large language models (LLMs) and humans as experts
to decide on edge directions.

@@

Since our approach relies on expert knowledge to determine edge orientations,
we explored the potential of using an LLM for this task. We applied our causal
discovery procedure using the Gemini 1.5 Flash model as the expert. We simulate
the \emph{weak ancestral oracle} using the LLM by asking it to choose the
causal direction between a pair of variables. The LLM is provided a description
of each of the variables for the task. We have provided the prompt used in
Appendix~\ref{section:llms}. 
Figure~\ref{fig:unexplained_ll} illustrates this distinction by comparing GES
to \textsc{ExpertInLoop} on the Adult income dataset across iterations of a
single run of the algorithms.
Figure~\ref{fig:adult_llm} shows the DAG learned
on the Adult Income dataset using the LLM as the expert. Compared to other
methods that use LLMs for causal discovery, our approach combines a data-driven
method with LLMs. We believe this should help in reducing the amount of
mistakes made by the LLM as well as reduce the amount of information needed
from the LLM.

@@

To enable researchers to easily apply our approach to their own datasets, we
developed an interactive web-based tool (Figure~\ref{fig:web}) for constructing
DAGs. Users can upload their dataset, which initializes an empty DAG with nodes
corresponding to the dataset's variables. They can then specify a p-value
threshold and a measure of association threshold. To accommodate mixed data
types, we use a measure of association based on canonical correlations
(detailed in Appendix~\ref{sec:mixed_association}).

Using the specified thresholds, the tool visually highlights variable pairs
with a residual association greater than the threshold, marking them with red
edges. The thickness of these edges represents the strength of association,
helping users prioritize which variable pairs to address first. Similarly, if
an existing edge is superfluous, it is highlighted in black, signaling
potentially unnecessary edge. Using this information, users can iteratively
modify the model by adding or removing edges. The tool also computes the Root
Mean Square Error of Approximation (RMSEA) to provide an estimate of the
overall fit of the model. Once satisfied with the constructed DAG, users can
export the model for further analysis.

\section{Conclusions}
\todo[inline]{Rewrite}
Researchers often prefer to construct DAGs manually rather than using automated
algorithms due to practical challenges in applying automated methods. To assist
in the process, we developed an iterative structure learning method that
integrates manual construction with data-driven insights. This approach bridges
the gap between fully manually and fully automated methods by providing a
systematic way to construct DAGs while utilizing data.

In future we would like to extend our approach to latent variables and Acyclic
Directed Mixed Graphs (ADMGs), explore the potential of combining our approach
with pairwise orientation methods. We would also improve the web interface for
better usability. 
\bibliography{references}

>>>>

\begin{figure}[t!]
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/unexplained_effect.pdf}
		\caption{}
	\end{subfigure}%
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/ll.pdf}
		\caption{}
	\end{subfigure}
	\caption{LLM-in-the-loop causal discovery on the Adult Income datases. As the ground 
truth is unknown, two measures of fit are shown over 30 iterative modifications: 
total residual association $\tau$ (a, lower is bettter); log-likelihood (a, higher is better).
	}
	\label{fig:unexplained_ll}
\end{figure}

@@

We now present two proof-of-concept implementations of our approach, 
geared towards using  large language models (LLMs) and humans as experts
to decide on edge directions. We use the Adult Income dataset from the 
introduction as an example. There is no known ground truth for this dataset,
and it contains mixed types of data; here, we use a version that contains
ordinal and nominal variables and use the RMSEA as a measure of association
to prioritize modifications (Section~\ref{sec:background}).

@@

Since our approach relies on expert knowledge to determine edge orientations,
we explored the potential of using an LLM for this task. We applied our causal
discovery procedure using the Gemini 1.5 Flash model as the expert. We simulate
the \emph{weak ancestral oracle} using the LLM by asking it to choose the
causal direction between a pair of variables. The LLM is provided a description
of each of the variables and given the prompt shown in
Appendix~\ref{section:llms}. 
Figure~\ref{fig:unexplained_ll} shows how the model fit to the data improves 
across 30 iterations compared to GES for a run of the algorithm.
Figure~\ref{fig:adult_llm} shows the DAG learned
on the Adult Income dataset using the LLM as the expert. We can see 
that the model fit improves comparably to a greedy approach and 
the resulting DAG contains largely sensible edge directions.

@@

To enable researchers to easily apply our approach to their own datasets, we
developed an interactive web-based tool (Figure~\ref{fig:web}) for constructing
DAGs. Users can upload their dataset, which initializes an empty DAG with nodes
corresponding to the dataset's variables. They can then specify a p-value
threshold and a threshold for a minimal association strength. The tool 
then visually highlights variable pairs
with a residual association greater than the threshold, marking them with red
edges; to avoid cluttering, no lines are drawn for residual associations that
are statistically insignificanct or below the specified threshold.
The thickness of these edges represents the strength of association,
helping users prioritize which variable pairs to address first; however, note that
users are free to choose which variables to connect. Similarly, if
an existing edge is found to be potentially superfluous (statistically insignificant),
it is highlighted in black. Using this information, users can iteratively
modify the model by adding or removing edges. The tool also computes the Root
Mean Square Error of Approximation (RMSEA) based on Shipley's C test, a global test 
of model fit based on the implied CIs \citep{Shipley2000}, to provide an estimate of the
overall fit of the model. Once satisfied with the constructed DAG, users can
export the model for further analysis.

\section{Conclusions}

Researchers often prefer to construct DAGs manually rather than using automated
algorithms due to practical challenges in applying automated methods. To assist
this process, we developed an iterative structure learning method that
integrates manual construction with data-driven feedback, bridging
the gap between fully manually and fully automated methods.
The idea of augmenting structure learning by domain knowledge is of course not 
new. Common ways to provide such background knowledge are "whitelists" or "blacklists"
of edges that must or must not be present in the resulting DAG, or specification
of a temporal structure between the variables. 
Compared to such approaches, where domain knowledge 
is specified and supplied beforehand, the novelty of our method lies in the interactive
back-and-forth between domain knowledge and data, which provably leads to the correct
result in a polynomial amount of steps (if the expert answers all
ancestral queries correctly).

Avenues for future work include extending this approach to latent variables 
as encoded by acyclic directed mixed graphs (ADMGs). Further, if experts make 
mistakes that lead to inconsistencies during model construction (such as cycles),
this information could be leveraged in a more systematic way to allow backtracking
and recovery.
\bibliography{references}
